Modeling the near-wake of a vertical-axis cross-flow turbine with 2-D and 3-D RANS

### 2016-03-19 paper.tex
Add results and conclusions from CFD chapter

Computations for the 3-D cases were run on 192 processes and took on the order
of 1,000 CPU hours per second of simulated time. The 2-D simulations were run on
a single processor and took on the order of one CPU hour per second.

@@

Predictions for both the mean rotor power and drag coefficients are shown in
Figure~\ref{fig:perf-comp}. In general, the 2-D CFD cases both overpredict
turbine loading, which is likely due to their increased blockage ratio,
unresolved blade end effects, and lack of blade support struts.

The 3-D simulations fair better at predicting the experimental measurements,
with the Spalart--Allmaras model performing relatively better. The apparent
overprediction of rotor drag coefficient could be an effect of the experimental
procedure, where the ``tare drag'' from the turbine mounting structure was
measured without a turbine installed, then subtracted in post-processing.
Technically the local flow field will have changed, and hence the tare drag will
change as well.

    \includegraphics[width=0.9\textwidth]{figures/perf_bar_chart}

@@

    \label{fig:perf-comp}
\end{figure}

@@

Visualizing the complicated wake generated by the turbine is presented for the
2-D and 3-D Spalart--Allmaras cases in Figure~\ref{fig:vorticity-3D} and
Figure~\ref{fig:vorticity-3D}, respectively. It can be seen how the upstream
blade---as it turns back into the streamwise direction---is shedding a large
amount of spanwise vorticity due to the separated flow. In the 3-D case, strong
tip vortices are also present, which trace the ``contracting'' wake flow on the
$-y$ side of the turbine associated with the induced vertical velocity field.
The 3-D dynamic stall vortex also shows assymmetry about the $x$--$y$ mid-rotor
plane; once again highlighting the importance of three-dimensional effects on
wake dynamics.

\begin{figure}
    \includegraphics[width=0.65\textwidth]{figures/2D_vorticity_SA_964}

@@

    \label{fig:vorticity-2D}
\end{figure}

@@

    \includegraphics[width=0.8\textwidth]{figures/3D_vorticity_SA_964_10-threshold}

@@

    \label{fig:vorticity-3D}
\end{figure}

Mean velocity profiles at one turbine diameter downstream are shown in
Figure~\ref{fig:profiles}. The 2-D results suffer from a blockage mismatch, i.e.
keeping the proximity of the walls constant increases the blockage ratio. The
3-D results, however, show good agreement with the experiments.

@@

    \includegraphics[width=0.95\textwidth]{figures/profiles}

@@

    \label{fig:profiles}
\end{figure}

Turbulence kinetic energy profiles are also shown in Figure~\ref{fig:profiles}.
The turbulence kinetic energy is calculated as
\begin{equation}
    k = k_{\mathrm{RA}} + \frac{1}{2} \left(
    \overline{U^\prime}^2 +
    \overline{V^\prime}^2 +
    \overline{W^\prime}^2 \right),
    \label{eq:k}
\end{equation}
where $U^\prime = U - \overline{U}$ and $k_{\mathrm{RA}}$ is the kinetic energy
calculated by the turbulence model, which is zero for the SA model. Note that
the statistics of the Reynolds-averaged velocity are calculated from a time
series that has been downsampled to 50 Hz. It is assumed that this is far enough
from the blade passage frequency that differences from the variance in the
original velocity will be negligible.

@@

In order to visualize the mean velocity field, vector arrows for the mean
cross-stream and vertical components are superimosed on top of contours of the
streamwise component at $X/D=1$ in Figure~\ref{fig:mean-velocity}. Both CFD
models predict the general structure of the mean velocity well, though the SA
case has slightly larger vertical mean flows, which could be due to stronger tip
vortex generation, or lower diffusivity compared with the SST model. Additional
discrepancies between CFD and experiments may be due to the top slip boundary
condition versus the experiment's free surface.

@@

        \includegraphics[clip, trim=0 0.1in 0 0.2in,
        width=0.75\textwidth]{figures/meancontquiv_exp}

        \caption{Mean velocity field at $x/D=1$ from experiments
            \cite{Bachant2015-RVAT-Re-dep-data}.}

        \label{fig:meancontquiv-exp}
    \end{subfigure}

@@

        \includegraphics[clip, trim=0 0.2in 0 0.15in,
        width=0.8\textwidth]{figures/meancontquiv_kOmegaSST}

@@

        \includegraphics[clip, trim=0 0.2in 0 0.15in,
        width=0.8\textwidth]{figures/meancontquiv_SpalartAllmaras}

@@

    \label{fig:mean-velocity}
\end{figure}

@@

Turbulence kinetic energy contours for the experimental measurements and each
CFD case at $x/D=1$ are presented in Figure~\ref{fig:kcont}. As seen in the
profiles in Figure\ref{fig:profiles}, the SA model is resolving very little of
the flow unsteadiness. In contrast, the SST model does a good job predicting the
locations and magnitudes of various peaks in $k$. These are generated along the
top of the turbine via tip vortex shedding, and the $-y$ side of the turbine via
dynamic stall. We do however see the smearing effect from the dynamic stall
vortex centered around $z/H=0$, which is likely more of an issue with wake
evolution rather than wake generation.

@@

        \includegraphics[width=0.95\textwidth]{figures/kcont_exp}

        \caption{Turbulence kinetic energy at $x/D=1$ from experiments
            \cite{Bachant2015-RVAT-Re-dep-data}.}

@@

        \includegraphics[width=0.95\textwidth]{figures/kcont_kOmegaSST}

@@

        \includegraphics[width=0.95\textwidth]{figures/kcont_SpalartAllmaras}

@@

    \label{fig:kcont}
\end{figure}

@@

To get an overall idea of the wake recovery predicted by each model, we
rearrange the streamwise component of the Navier--Stokes equation to isolate
$\partial U / \partial x$---following Bachant and
Wosnik\cite{Bachant2015-JoT}---and compute each term at $X/D = 1$ to compare
with the experimental results.

@@

Normalized weighted averages for each recovery term at $x/D=1$ are computed and
multiplied by the cross-sectional area of the measurement plane, or the channel
width in the 2-D cases. Results are shown in a bar chart in
Figure~\ref{fig:recovery}. Consistent with the relatively large Reynolds number
regime, viscous transport is essentially negligible compared with other
mechanisms. Cross-stream advection---or the tendency of streamlines to diverge
and reduce the streamwise momentum---produces a negative effect for all cases,
though the 3-D SST model predicts significantly lower values. Vertical advection
is by definition zero for the 2-D cases. The 3-D cases show varying
results---with the SST model overpredicting and SA underpredicting the vertical
velocity's effect on replenishing streamwise momentum.

Turbulent transport and streamwise pressure gradient terms show the largest
discrepancy between results. The 3-D SST case, despite doing a good job
predicting turbulence kinetic energy, significantly overpredicts the turbulent
transport term, while other CFD cases are comparable with the experiment. This
seems to be balanced by a large adverse pressure gradient, which is also present
to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
cases create a wake where the pressure gradient is acting to accelerate the flow
at $x/D=1$. Unfortunately, pressure data were not acquired from the experiment,
though the 3-D delayed detached eddy simulation (DDES) of Boudreau and Dumas
\cite{Boudreau2015} concur with the adverse pressure gradient condition.

    \includegraphics[width=0.9\textwidth]{figures/mom_bar_graph}

    \caption{Weighted sum normalized momentum recovery terms for each CFD case
        and experiments\cite{Bachant2015-RVAT-Re-dep-data} at $x/D=1$.}

    \label{fig:recovery}
\end{figure}

@@

A cross-flow turbine was modeled using the $k$--$\omega$ SST and
Spalart--Allmaras Reynolds-averaged Navier--Stokes turbulence models to test
their abilities to predict turbine performance and near-wake dynamics. It was
observed that when modeled in 2-D, the performance is over-predicted compared to
the data from the tow tank experiments, which was expected due to omission of
blade end effects, support strut drag, and increased blockage. Vertical (or
axial) wake dynamics were unresolved in the 2-D model, despite being identified
as a significant contributor to streamwise wake recovery, which casts doubt on
the 2-D model's applicability as a tool to study array spacing effects.

These results may also help develop new tip loss corrections for blade element
type models, which currently only exist for axial-flow rotors.

>>>>

Turbine operation in all cases was simulated for 10 seconds, or approximately
six rotor revolutions. Computations for the 3-D cases were run on 192 processes
(24 nodes $\times$ 8 cores each) and took on the order of 1,000 CPU hours per
second of simulated time. The 2-D simulations were run on a single processor and
took on the order of one CPU hour per second.

Flow statistics were calculated over the time interval spanning $5$--$10$
seconds, or approximately three rotor revolutions, from time series downsampled
to 50 Hz. It is assumed that the downsampling frequency is high enough above the
blade passage frequency such that differences from the variance (for computing
turbulence statistics) in the original velocity will be negligible.

@@

Mean performance quantities---power and overall rotor drag coefficients---were
calculated for the interval after two full rotor revolutions, beyond which
values had reached an approximately periodic state. Predictions for both the
mean rotor power and drag coefficients are shown in
Figure~\ref{fig:br-cfd-perf-bar-chart}. In general, the 2-D CFD cases both
significantly overpredict turbine loading and therefore mechanical power output,
which is likely due to their increased blockage ratio, unresolved blade end
effects, and lack of blade support struts.

The 3-D simulations fair better at predicting the experimental measurements. The
Spalart--Allmaras model's mean power coefficient was 0.24 while the SST's was
0.33, which represent a 6\% under- and 30\% overprediction, respectively,
compared with the experiments. The apparent overprediction of rotor drag
coefficient could be an effect of the experimental procedure, where the ``tare
drag'' from the turbine mounting structure was measured without a turbine
installed, then subtracted in post-processing. Technically the local flow field
will have changed, and hence the tare drag will change as well.

    \includegraphics[width=0.95\textwidth]{BR-CFD_perf_bar_chart}

@@

    \label{fig:br-cfd-perf-bar-chart}
\end{figure}

@@

Visualizations of the complex vorticity field generated by the turbine are
presented for the 2-D and 3-D Spalart--Allmaras cases in
Figure~\ref{fig:br-vorticity-3d} and Figure~\ref{fig:br-vorticity-3d},
respectively. It can be seen how the upstream blade---as it turns back into the
streamwise direction---is shedding a large amount of spanwise vorticity due to
the separated flow. In the 3-D case, strong tip vortices are also present, which
trace the ``contracting'' wake flow on the $-y$ side of the turbine associated
with the induced vertical velocity field. The 3-D dynamic stall vortex also
shows asymmetry about the $x$--$y$ mid-rotor plane; once again highlighting the
importance of three-dimensional effects on wake dynamics.

    \includegraphics[width=0.75\textwidth]{BR-CFD_2D_vorticity_SA_964}

@@

    \label{fig:br-vorticity-2d}
\end{figure}

@@

    \includegraphics[width=0.75\textwidth]{BR-CFD_3D_vorticity_SA_964_10-threshold}

@@

    \label{fig:br-vorticity-3d}
\end{figure}

Mean velocity profiles at one turbine diameter downstream are shown in
Figure~\ref{fig:br-cfd-profiles}. The 2-D results suffer from a blockage
mismatch, i.e., keeping the proximity of the walls constant increases the
blockage ratio. The 3-D results, however, show good agreement with the
experiments.

@@

    \includegraphics[width=0.95\textwidth]{BR-CFD_profiles}

@@

    \label{fig:br-cfd-profiles}
\end{figure}

Turbulence kinetic energy profiles are also shown in
Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy is calculated as
\begin{equation}
k = k_{\mathrm{RA}} + \frac{1}{2} \left(
\overline{U^\prime}^2 +
\overline{V^\prime}^2 +
\overline{W^\prime}^2 \right),
\label{eq:k}
\end{equation}
where $U^\prime = U - \overline{U}$ and $k_{\mathrm{RA}}$ is the kinetic energy
calculated by the turbulence model, which is zero for the SA model.

@@

In order to visualize the mean velocity field, vector arrows for the mean
cross-stream and vertical components are superimosed on top of contours of the
streamwise component at $X/D=1$ in Figure~\ref{fig:br-cfd-mean-velocity}. Both
CFD models predict the general structure of the mean velocity well, though the
SA case has slightly larger vertical mean flows, which could be due to stronger
tip vortex generation, or lower diffusivity compared with the SST model.
Additional discrepancies between CFD and experiments may be due to the top slip
boundary condition versus the experiment's free surface.

@@

        \includegraphics[clip, trim=0 0.1in 0.3in 0.2in,
        width=0.72\textwidth]{RVAT-Re-dep_meancontquiv_10}

        \caption{Mean velocity field at $x/D=1$ from experiments
            \cite{Bachant2016-RVAT-Re-dep}.}

        \label{fig:br-cfd-meancontquiv-exp}
    \end{subfigure}

@@

        \includegraphics[clip, trim=0 0.2in 0 0.15in,
        width=0.8\textwidth]{BR-CFD_meancontquiv_kOmegaSST}

@@

        \includegraphics[clip, trim=0 0.2in 0 0.15in,
        width=0.8\textwidth]{BR-CFD_meancontquiv_SpalartAllmaras}

@@

    \label{fig:br-cfd-mean-velocity}
\end{figure}

@@

Turbulence kinetic energy contours for the experimental measurements and each
CFD case at $x/D=1$ are presented in Figure~\ref{fig:br-cfd-kcont}. As seen in
the profiles in Figure~\ref{fig:br-cfd-profiles}, the SA model is resolving very
little of the flow unsteadiness. In contrast, the SST model does a good job
predicting the locations and magnitudes of various peaks in $k$. These are
generated along the top of the turbine via tip vortex shedding, and the $-y$
side of the turbine via dynamic stall. We do however see the smearing effect
from the dynamic stall vortex centered around $z/H=0$, which is likely more of
an issue with wake evolution rather than wake generation.

@@

        \includegraphics[width=0.95\textwidth]{RVAT-Re-dep_k_contours_10}

        \caption{Turbulence kinetic energy at $x/D=1$ from experiments
            \cite{Bachant2016-RVAT-Re-dep}.}

@@

        \includegraphics[width=0.95\textwidth]{BR-CFD_kcont_kOmegaSST}

@@

        \includegraphics[width=0.95\textwidth]{BR-CFD_kcont_SpalartAllmaras}

@@

    \label{fig:br-cfd-kcont}
\end{figure}

@@

To get an overall idea of the wake recovery predicted by each model, we
rearrange the streamwise component of the Navier--Stokes equation to isolate
$\partial U / \partial x$---following Bachant and
Wosnik~\cite{Bachant2015-JoT}---and compute each term at $X/D = 1$ to compare
with the experimental results.

@@

Normalized weighted averages for each recovery term at $x/D=1$ are computed and
multiplied by the cross-sectional area of the measurement plane, or the channel
width in the 2-D cases. Results are shown in a bar chart in
Figure~\ref{fig:br-cfd-recovery}. Consistent with the relatively large Reynolds
number regime, viscous transport is essentially negligible compared with other
mechanisms. Cross-stream advection---or the tendency of streamlines to diverge
and reduce the streamwise momentum---produces a negative effect for all cases,
though the 3-D SST model predicts significantly lower values. Vertical advection
is by definition zero for the 2-D cases. The 3-D cases show varying
results---with the SST model overpredicting and SA underpredicting the vertical
velocity's effect on replenishing streamwise momentum.

Turbulent transport and streamwise pressure gradient terms show the largest
discrepancy between results. The 3-D SST case, despite doing a good job
predicting turbulence kinetic energy, significantly overpredicts the turbulent
transport term, while other CFD cases are comparable with the experiment. This
seems to be balanced by a large adverse pressure gradient, which is also present
to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
cases create a wake where the pressure gradient is acting to slightly accelerate
the flow at $x/D=1$. Unfortunately, pressure data were not acquired from the
experiment, though the 3-D delayed detached eddy simulation (DDES) of Boudreau
and Dumas \cite{Boudreau2015} concur with the adverse pressure gradient
condition.

    \includegraphics[width=0.9\textwidth]{BR-CFD_mom_bar_graph}

    \caption{Weighted sum normalized momentum recovery terms for each CFD case
        and experiments\cite{Bachant2016-RVAT-Re-dep} at $x/D=1$.}

    \label{fig:br-cfd-recovery}
\end{figure}

@@

A cross-flow turbine was modeled using the $k$--$\omega$ SST and
Spalart--Allmaras (SA) Reynolds-averaged Navier--Stokes turbulence models to test
their abilities to predict turbine performance and near-wake dynamics. It was
observed that when modeled in 2-D, the performance is over-predicted compared to
the data from the tow tank experiments, which was expected due to omission of
blade end effects, support strut drag, and increased blockage. Vertical (or
axial) wake dynamics were unresolved in the 2-D model, despite being identified
as a significant contributor to streamwise wake recovery, which casts doubt on
the 2-D model's applicability as a tool to study array spacing effects.

The 3-D blade-resolved RANS simulations predicted turbine performance and
near-wake quite well, where the SA model results were closest to the
experimentally measured performance. The SA model's effectiveness gives hope for
the prospect of replacing some physical modeling with blade-resolved CFD.
However, the  overprediction of power coefficient by the SST model highlights
the level of technological and economic risk involved with CFD. That is to say,
the SST model, despite being considered relatively trustworthy in these flow
conditions, would overestimate power output by 30\%.

Both the SST and SA models did a good job predicting the mean velocity field at
one turbine diameter downstream, but the SST model was more effective at
predicting turbulence kinetic energy, since it is solved for in the turbulence
model equations. Streamwise momentum recovery terms were computed from the CFD
results over an entire cross-section of the domain at $x/D=1$. Values for
transport due to the mean pressure gradient and turbulent fluctuations varied a
lot between the two turbulence models. Both models, however, we able to at least
qualitatively resolve the vertical velocity field, which will be crucial to
predicting the performance of closely spaced CFTs.

The effectiveness at predicting the mean velocity field gives credibility to the
prospect of using the computed flow field to interpolate the experimental
results, such that the CFD results can be used as a target for a low-order wake
generator or force parameterization. These results may also help develop new tip
loss corrections for blade element type models, which currently only exist for
axial-flow rotors, since they provide access to the pressure and shear forces
over the entire blade surface, which were not measured experimentally.
Ultimately, however, the computational cost of 3-D simulations---about 10,000
CPU hours per operating case---may be too expensive to be used for CFT
engineering work, especially considering the uncertainty involved compared with
physical model studies. Since the 3-D wake results appeared to be only weakly
asymmetrical in the $x$--$y$ plane about $z/H=0$, it may be possible to reduce
computing power by only modeling the top half of the rotor and using a symmetry
boundary condition at the mid-plane. However, 3-D blade-resolved RANS will
likely not be practical for turbine array simulations for quite some time, and
until they are, low-order models will need to be developed and improved for this
purpose.
